---
title: "Parameter Estimation"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2026 | University of Portland"
output:
  slidy_presentation:
    font_adjustment: +2
    footer: "| MTH-361A Spring 2026 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gghighlight)
library(gridExtra)
library(latex2exp)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Introduce confidence intervals**
- **Know how to compute confidence intervals**
- **Develop an understanding of interpreting confidence intervals**
::::

:::: {.column width=15%}
::::

## Confidence Intervals

A plausible range of values for the population parameter is called a *confidence interval*.

:::: {.column width=49%}
**Fishing analogy:**

A confidence interval is like using a fishing net, rather than a spear, to catch fish in a murky lake.

* *Parameter (the fish):* The true population parameter (e.g., mean fish length). This is assumed to be fixed but unknown.
* *Sample statistic (the catch):* The observation (e.g., the data collected from the lake) and the point estimate (e.g., sample mean fish length).
::::

:::: {.column width=49%}
**The interval:**

We need to set the size of the fishing net first before casting it into the lake.

* *Confidence level (the net size):* A higher confidence level (e.g., 99\%) means a wider, larger net, increasing the chance of catching the fish.
* *Confidence interval (the net):* The range of values calculated from the sample.
::::

::: {style="color: red;"}
$\star$ If we report a point estimate, we probably won't hit the exact population parameter. If we report a range of plausible values we have a good shot at capturing the parameter.
:::

## Spam Emails

Suppose we want to estimate the *number* of spam emails of an account.

:::: {.column width=49%}
**Data summary:**

```{r}
set.seed(42)
n <- 200
email_sub <- email %>% 
  slice_sample(n = n) %>% 
  group_by(spam) %>% 
  summarise(count = n(),
            prop = count/n)
```

| Not Spam | Spam |
|:---:|:---:|
| 184 | 16 |

* The data shows that there are $16$ spam emails out of $200$ emails.
* The population of interest is all emails of one account.
* The data we have is a random sample from the population.
::::

:::: {.column width=49%}
**Point estimate:**

* The *sample statistic* is $$\overline{x} = 16,$$ which is our *point estimate*.
* $\overline{x}$ is the number of spam emails, an estimate to an unknown *population parameter*.

::: {style="color: red;"}
$\star$ For the purpose of this example, the sample mean notation $\overline{x}$ can be thought of as the mean of a binomial r.v..
:::
::::

## Sampling Distribution

The point estimate for the number of spam emails is $\overline{x} = 16$ and the sample size is $n = 200$.

:::: {.column width=49%}
**Assumptions:**

* The emails are independently sampled from the same account
* The probability of getting a randomly selected spam email is fixed.
* We assume that CLT conditions are satisfied.
* The sampling distribution of the number of spam emails exhibits a binomial distribution.
::::

:::: {.column width=49%}
**Sampling distribution of $\overline{x}$:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# set parameters, outcomes, and probabilities
n <- 200 # number of trials
p <- 16/n # set "success" probability
x_outcomes <- seq(0,31,1) # set outcome from 0 to n
binom_pmf <- dbinom(x_outcomes,n,p) # function `dbinom()` is the Binomial PMF 
binom_ev <- n*p # expected value
df <- tibble(k=x_outcomes, pmf=binom_pmf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=k,y=pmf)) + 
  geom_point(size=3,color="black") + # size here is defined for all points
  geom_segment(aes(x=k,xend=k, # draws a line between two defined points
                   y=rep(0,length(k)),yend=pmf), 
                   color="black") + 
  geom_vline(xintercept = n*p, color="red", linetype="dashed") + 
  xlab("") + 
  ylab("probability") + 
  ylim(0,0.12) + 
  scale_x_discrete(limits=c(16),label=c(TeX("$\\bar{x} = 16$"))) + 
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ The domain of this distribution is $0 \le x \le 200$ (the graph is truncated) because there are $n = 200$ samples and the $x$ value is the number of spam emails out of $n$, where it could be $0$, $200$, or in between.
:::
::::

## Confidence Interval of the Number of Spam Emails

The confidence level is a probability that we set. This is how wide our interval is going to be.

:::: {.column width=49%}
**Confidence Level:**

* Suppose we want a "$90$\% confidence interval". So, the confidence level is $0.90$.
* This is the interval probability of the sampling distribution.
* The goal is to find $a$ and $b$, so that $$P(a \le \overline{X} \le b) \approx 0.90,$$ where $X$ is a binomial r.v..
* The middle probability is $0.90$ and the tail probabilities sums to $0.10$, with individual tails to be $0.05$, assuming symmetry.
* Probability components:

  - lower bound: $\displaystyle P(\overline{X} \le a) \approx 0.05 \longrightarrow a = 10$
  - upper bound: $\displaystyle P(\overline{X} \le b) \approx 0.95 \longrightarrow b = 23$
::::

:::: {.column width=49%}
**Confidence Interval:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# set parameters, outcomes, and probabilities
n <- 200 # number of trials
p <- 16/n # set "success" probability
x_outcomes <- seq(0,31,1) # set outcome from 0 to n
binom_pmf <- dbinom(x_outcomes,n,p) # function `dbinom()` is the Binomial PMF 
binom_ev <- n*p # expected value
df <- tibble(k=x_outcomes, pmf=binom_pmf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=k,y=pmf)) + 
  geom_point(size=3,color="blue") + # size here is defined for all points
  geom_segment(aes(x=k,xend=k, # draws a line between two defined points
                   y=rep(0,length(k)),yend=pmf), 
                   color="blue") + 
  gghighlight(k>=10&k<=23, unhighlighted_colour = "black") + 
  geom_vline(xintercept = n*p, color="red", linetype="dashed") + 
  geom_vline(xintercept = 10, color="red") + 
  geom_vline(xintercept = 23, color="red") + 
  xlab("") + 
  ylab("probability") + 
  ylim(0,0.12) + 
  scale_x_discrete(limits=c(10,16,23),label=c(TeX("$10$"),TeX("$16$"),TeX("$23$"))) + 
  theme_minimal() # set theme of entire plot

# display plot
p1
```


$$10 \le \overline{x} \le 23$$

::: {style="color: red;"}
$\star$ Interpreting this interval requires you to understand how CLT works and a basic understanding of the frequentist interpretation of probability.
:::

**Using R:**

```{r echo = TRUE}
cl <- 0.90 # confidence level
cl_tail <- 1-cl # tail probabilities
lb <- qbinom(cl_tail/2,200,16/200) # lower bound
ub <- qbinom(cl+(cl_tail/2),200,16/200) # upper bound
c(lb,ub) # confidence interval as an ordered list
```
::::

## Normal Approximation

We have $16$ spam emails and $184$ not spam emails, both are more than $10$, and the sample size $n = 200$ is considered large enough. This *rule of thumb* says that we can approximate the binomial using the normal distribution.

:::: {.column width=49%}
**Normal approximation:**

* The normal approximation to the binomial would yield:
$$
\begin{aligned}
\overline{x} & = n\hat{p} \longleftarrow \text{ sample mean} \\
s^2 & = n\hat{p}(1-\hat{p}) \longleftarrow \text{sample variance}
\end{aligned}
$$
* Note that $$\hat{p} = \frac{\text{number of spam emails}}{n}.$$ This is known as the sample proportion.
* So, $\displaystyle \hat{p} = \frac{16}{200} = 0.08$.
::::

:::: {.column width=49%}
**Confidence interval assuming normality:**

$$
\overline{x} \pm z^* \cdot SE
$$
where $SE$ is called the *standard error*. Here, $\displaystyle SE = \sqrt{n\hat{p}(1-\hat{p})}$. The term $z^*$ is called the critical value, which can be computed using R.

So, the interval is 

$$
\begin{aligned}
\overline{x} & \pm z^* \cdot \sqrt{n\hat{p}(1-\hat{p})} \\
16 & \pm 1.645\left(\sqrt{200(0.08)(1-0.08)})\right)
\end{aligned}
$$

$$9.689 \le \overline{x} \le 22.311$$

**Using R:**

```{r echo=TRUE}
cl <- 0.90 # confidence level
cl_tail <- 1-cl # tail probabilities
lb <- qnorm(cl_tail/2,16,sqrt(200*0.08*(1-0.08))) # lower bound
ub <- qnorm(cl+(cl_tail/2),16,sqrt(200*0.08*(1-0.08))) # upper bound
c(lb,ub) # confidence interval as an ordered list
```
::::

## Facebook’s categorization of user interests

Most commercial websites (e.g. social media platforms, news out- lets, online retailers) collect a data about their users’ behaviors and use these data to deliver targeted content, recommendations, and ads.

To understand whether Americans think their lives line up with how the algorithm-driven classification systems categorizes them, Pew Research asked **a representative sample of 850 American Facebook users** how accurately they feel the list of categories Facebook has listed for them on the page of their supposed interests actually represents them and their interests. **67% of the respondents said that the listed categories were accurate**.

Estimate the true proportion of American Facebook users who think the Facebook categorizes their interests accurately.

## Point Estimate and Standard Error

The goal of parameter estimation is to find a range of possible values (confidence interval).

**Given information**

* $\hat{p} = 0.67 \longleftarrow \text{point estimate}$
* $n = 850 \longleftarrow \text{sample size}$
    - The expected number of users who think the Facebook categorizes their interests accurately is $850 \times 0.67 \approx 569.5$ (569 or 570).
    - There around 280.5 (280 or 281) users think the opposite.
* Let $p$ bet he true population proportion and $\hat{p}$ be the sample proportion.
    
**The Confidence Interval**

We want to find the 95% confidence interval using the formula: $$\text{point estimate} \pm 1.96 \times \text{SE}$$ where **SE is the standard error**.

This can be written as 
\[
\begin{aligned}
0.67 & \pm 1.96 \times \sqrt{\frac{0.67 (1-0.67)}{850}} \\
0.67 & \pm 1.96 \times 0.0161 \\
& \longrightarrow (0.67-0.0316,0.67+0.0316) \\
& \longrightarrow (0.6384,0.7016)
\end{aligned}
\]

Thus, the 95% interval for estimating the true $p$ is between 0.6384 and 0.7016.

## Interpretation

Which of the following is the correct interpretation of this confidence interval? We are 95% confident that:

i. 64% to 67% of American Facebook users in this sample think Facebook categorizes their interests accurately.
ii. 64% to 67% of all American Facebook users think Facebook categorizes their interests accurately
iii. there is a 64% to 67% chance that a randomly chosen American Facebook user’s interests are categorized accurately.
vi. there is a 64% to 67% chance that 95% of American Facebook users’ interests are categorized accurately.

::: {style="color: red;"}
$\star$ 64% to 67% of all American Facebook users think Facebook categorizes their interests accurately.
:::

::: {style="color: blue;"}
$\dagger$ Why do we interpret the confidence interval this way?
:::

## What does 95% Confident Mean?

Suppose we took many samples and built a confidence interval from each sample using the equation $$\text{point estimate} \pm 1.96 \times \text{standard error}.$$

Then about 95% of those intervals would contain the true population proportion ($p$).

## Width of an interval

If we want to be more certain that we capture the population parameter, i.e. increase our confidence level, should we use a wider interval or a smaller interval?

::: {style="color: red;"}
$\star$ A wider interval.
:::

Can you see any drawbacks to using a wider interval?

::: {style="color: red;"}
$\star$ If the interval is too wide it may not be very informative.
:::

## Changing the Confidence Level

$$\text{point estimate} \pm z^{\star} \times \text{SE}.$$

* In a confidence interval, $z^{\star} \times \text{SE}$ is called the **margin of error**, and for a given sample, the margin of error changes as the confidence level changes.
* In order to change the confidence level we need to adjust $z^{\star}$ in the above formula.
* Commonly used confidence levels in practice are 90%, 95%, 98%, and 99%.
* For a 95% confidence interval, $z^{\star} = 1.96$.
* However, using the standard normal distribution, it is possible to find the appropriate $z^{\star}$ for any confidence level.

## 95% Confidence Interval

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(z=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=z,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_ribbon(data=subset(df_norm,z>=-1.96 & z<=1.96),aes(x=z,ymax=dnorm(z,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ylab("density") + 
  ggtitle("0.95 Probability is Between -1.96 to 1.96") + # sets the title of the plot
  scale_x_discrete(limits=c(-1.96,0,1.96)) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

## 99.7% Confidence Interval

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(z=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=z,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_ribbon(data=subset(df_norm,z>=-2.97 & z<=2.97),aes(x=z,ymax=dnorm(z,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ylab("density") + 
  ggtitle("0.997 Probability is Between -2.97 to 2.97") + # sets the title of the plot
  scale_x_discrete(limits=c(-2.97,0,2.97)) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

## Finding $z^{\star}$ Exactly

Find the $z^{\star}$ for a 92\% confidence level.

:::: {.column width=50%}
**Process:**

* Confidence level is 0.92.
* Lower tail of the $Z \sim N(0,1)$ is $\frac{(1-0.92)}{2} = 0.04$.
* We want to find the $z$ score that would yield a 0.04 probability.
* Use the `qnorm()` function in R.

**Using R:**

```{r echo=TRUE}
cl <- 0.92 # confidence level
lt <- (1-cl)/2 # lower tail probability
qnorm(lt,0,1) # computes the z star
```
::::

:::: {.column width=49%}
```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='100%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(z=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=z,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_ribbon(data=subset(df_norm,z<=-1.75),aes(x=z,ymax=dnorm(z,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("0.04 Probability is up to -1.75") + # sets the title of the plot
  scale_x_discrete(limits=c(-1.75)) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='100%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(z=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=z,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_ribbon(data=subset(df_norm,z>=-1.75 & z<=1.75),aes(x=z,ymax=dnorm(z,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("0.92 Probability is Between -1.75 to 1.75") + # sets the title of the plot
  scale_x_discrete(limits=c(-1.75,0,1.75)) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```
::::
