---
title: "Orientation"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2026 | University of Portland"
output:
  slidy_presentation:
    font_adjustment: +5
    footer: "| MTH-361A Spring 2026 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
bibliography: ../../references.bib
csl: ../../apa.csl
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Communicate highlights of the Syllabus**
::::

:::: {.column width=15%}
::::

## Course Information

:::: {.column width=5%}
::::

:::: {.column width=45%}
- **Number:** MTH-361
- **Title:** Applied Statistics I
- **Section:** A
- **Term:** Spring 2026
- **Credits:** 3
:::: 

:::: {.column width=45%}
- **Instructor:** Dr. Quijano
- **Schedule:** TuTh 11:20 AM - 12:45 PM
- **Location** Franz Hall 217
::::

:::: {.column width=5%}
::::

## About the Instructor

### Dr. [Alex John Quijano](https://www.alexjohnquijano.com){target="_blank"}

:::: {.column width=1%}
::::

:::: {.column width=40%}
- **Office:** Buckley Center 279
- **Email:** quijano@up.edu
- **Hobbies and Interests:** Movies, anime, musicals, coffee 
- **Other Facts:** Two cats (orange and black)
::::

:::: {.column width=57%}
- **Education:** Applied Mathematics
- **Profession**: Assistant Professor of Mathematics
- **Teaching:** Probability and statistics, data science, linear algebra, calculus, differential equations
- **Research:** Mathematical interdisciplinary research in data science, modeling of natural languages, social and biological systems
::::

:::: {.column width=1%}
::::

## Modes of Communication

:::: {.column width=1%}
::::

:::: {.column width=48%}
### **Teams**

We will be using Teams as the main real-time communication tool for general announcements, question-answering discussions, and direct messages.

> The Teams for this course is [MTH-361A-sp26](https://teams.microsoft.com/l/team/19%3ALujyzbarzFjlZszUD1znX1UgPPd7aRVSummQ1WtAzYM1%40thread.tacv2/conversations?groupId=08d5ebee-6e90-47f3-80b4-fadc91987b08&tenantId=ea8f3949-231c-40b6-a33f-56873af96f87){target="_blank"}. 
::::

:::: {.column width=48%}
### **Class Website**

The syllabus, tentative topics schedule, and other information are posted on the course website.

> You can access the course website at [mth-361a-sp26](https://upmathematics.github.io/mth-361a-sp26/){target="_blank"}.

*Note that the course website can be accessed in Teams.*
::::

:::: {.column width=1%}
::::

## Structure and Expectations

:::: {.column width=5%}
::::

:::: {.column width=45%}
### **Class Structure**

  - Pre-reading
  - Short lecture
  - Worksheet activity
  - Group presentations or discussions
::::

:::: {.column width=45%}
### **Assignments**

  - Worksheets
  - Homework
  - Project Phases
::::

:::: {.column width=5%}
::::

## Assignments

### **Worksheets**

The purpose of the worksheets is for in-class group work and activities.

### **Homework**

The purpose of the homework is to provide practice with problems, and it will include tasks that involve R.

### **Project Phases**

The purpose of the project phases is to break the semester-long project into smaller tasks, allowing for incremental progress toward its completion.

## Submission Guidelines

### **Moodle**

We will be using the course [Moodle](https://learning.up.edu){target="_blank"} page for submitting assignments and giving feedback.

> The Moodle for this course is [MTH-361A - sp26 - Applied Statistics I](https://learning.up.edu/course/view.php?id=44593){target="_blank"}. You are already listed in Moodle for this course using your UP account.

## Resources

### **Textbook**

This course uses the following textbook - which is free to access - for reading and practice.

```{r echo=FALSE}
refs <- bibtex::read.bib("../../references.bib")
```

  * `r capture.output(print(refs["speegle2021probability"]))`
  
*Note that the textbook can be accessed in Teams.*
  
## Tools

### **R**

We will be using R, as a computational tool, for some assignments. These tools are free and open-source. 

> We will use the [Posit Cloud](https://login.posit.cloud/login?redirect=%2F){target="_blank"} service, an online computing environment.

*We will discuss more about this tool next time.*

## Academic Support

### **Help Hours**

> Dr. Alex John Quijano
> 
> * 4:00 PM - 5:00 PM Mondays to Fridays, Buckley Center 279
> * [One-to-One, Buckley Center 279](https://outlook.office.com/bookwithme/user/563068e7b8f84361b7e7eaf55bc932d8@up.edu/meetingtype/KXqoHwKuE0qdlfS1AiumEg2?bookingcode=31b921a7-10ad-4113-895c-50152bee77c0&anonymous&ismsaljsauthenabled&ep=mLinkFromTile){target="_blank"} <br>*Click on the link to sign-up for a 15-minute session.*

My walk-in help hours start week 2.

### **Math Resource Center**

The [Math Resource Center](https://www.up.edu/learningcommons/tutoring-services/math-resource-center/index.html){target="_blank"} offers walk-in tutoring and appointment-based tutoring.
