---
title: "Summarise Data by Category"
subtitle: "Applied Statistics"
author: "<span style='color:blue'>Solutions</span>"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

- You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
- Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

- For the exercises, please provide your answers in full sentences.
- Write your answers by replacing the text "[Write your answer here]".
- Whenever a problem asks you to run something in R, create an R code chunk and write your R commands inside that chunk.
- If a problem requires modifying an existing R code chunk, make the necessary changes at the appropriate lines within the given code.
- Please read the problems fully and carefully before answering.
- When finished, knit this `.Rmd` to `.html`.
- Submit this `.Rmd` and the recently knitted `.html` to Moodle.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
```

# Collaborators

::: {style="color: blue;"}
<span style='color:blue'>Solutions</span>
:::

# Exercises

## Problem 1

Create a code chunk. In this chunk, load the `palmerpenguins` package. These packages are pre-installed.

::: {style="color: blue;"}
```{r}
library(palmerpenguins)
```
:::

## Problem 2

Examine the `penguins` tibble. There are four categorical variables. What are they and what are the unique categories of each variable? Create a code chunk that uses the `distinct()` verb on the `penguins` tibble with the piping operator `%>%`.

::: {style="color: blue;"}
```{r}
penguins
```

The four categorical variables are `species`, `island`, `sex`, and `year`.

```{r}
penguins %>% 
  distinct(species)
```

```{r}
penguins %>% 
  distinct(island)
```

```{r}
penguins %>% 
  distinct(sex)
```

```{r}
penguins %>% 
  distinct(year)
```
:::

## Problem 3

Create a code chunk and use the verbs `group_by()` and `summarise()` to calculate the mean (`mean()`) and median (`median()`) body mass of each species in the `penguins` tibble. Make sure you use the piping operator `%>%` for readable and efficient data processing.

::: {style="color: blue;"}
```{r}
penguins %>% 
  group_by(species) %>% 
  summarise(mean_bm = mean(body_mass_g),
            median_bm = median(body_mass_g))
```

Some of the species mean and median `body_mass_g` appears to be `NA` values. This is because some rows are `NA` values when computing the statistics.
:::
