---
title: "Exponential and Normal Distributions"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2026 | University of Portland"
output:
  slidy_presentation:
    font_adjustment: +2
    footer: "| MTH-361A Spring 2026 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gghighlight)
library(latex2exp)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- 
- 
- 
::::

:::: {.column width=15%}
::::

## Visualizing The Exponential Distribution (1/2)

:::: {.column width=50%}
**Exponential Probability Density Function (PDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# exponential pdf
mean <- 15
lambda <- 1/mean
x_exp <- seq(0,60,0.10)
exp_pdf <- dexp(x_exp,lambda)

# convert pdf to tibble
df_exp <- tibble(x=x_exp,pdf=exp_pdf)

# plot the exponential distribution
ggplot(df_exp,aes(x=x,y=pdf)) + 
  geom_line(color="black",linewidth=1) + 
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda$","=","$\\frac{1}{",mean,"})$"))) + # sets the title of the plot
  theme_minimal()
```

::: {style="color: red;"}
$\star$ The exponential r.v. has infinitely sized sample space. The plot of the exponential PDF only shows the range $0 \le x \le 60$.
:::
::::

:::: {.column width=49%}
**Exponential R.V. and PDF:**

Let $\lambda$ be the rate parameter.

* $\displaystyle X \sim \text{Exp}\left(\lambda\right)$
* $\displaystyle f(x) = \lambda e^{-\lambda x}, \quad x \ge 0$

**Relevant R Functions:**

* `dexp` $\leftarrow$ PDF
* `rexp` $\leftarrow$ random sampling simulation
::::

## Visualizing The Exponential Distribution (2/2)

:::: {.column width=50%}
**Exponential Cumulative Distribution Function (CDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 15
lambda <- 1/mean
x_exp <- seq(0,60,0.10)
exp_cdf <- pexp(x_exp,lambda)

# convert cdf to tibble
df_exp <- tibble(x=x_exp,cdf=exp_cdf)

# plot the exponential distribution
ggplot(df_exp,aes(x=x,y=cdf)) + 
  geom_line(color="black",linewidth=1) + 
  ggtitle(TeX(paste("Exponential CDF ","$(\\lambda$","=","$\\frac{1}{",mean,"})$"))) + # sets the title of the plot
  theme_minimal()
```

::: {style="color: red;"}
$\star$ The CDF, in general, computes the cumulative probability of a given PDF.
:::
::::

:::: {.column width=49%}
**Exponential PDF and CDF:**

Let $\lambda$ be the rate parameter.

* $\displaystyle f(x) = \lambda e^{-\lambda x}, \quad x \ge 0$
* $\displaystyle P(X \le x) = \int_{0}^{x} f(t) \ dt, \quad x \ge 0$

**Relevant R Functions:**

* `pexp` $\leftarrow$ CDF
* `qexp` $\leftarrow$ Inverse CDF
::::


