---
title: "Exponential and Normal Distributions"
subtitle: "Applied Statistics"
author: "MTH-361A | Spring 2026 | University of Portland"
output:
  slidy_presentation:
    font_adjustment: +2
    footer: "| MTH-361A Spring 2026 | <a href='../../index.html'>Back to the Course Website</a>"
    css: ../_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r echo=FALSE, eval=TRUE,message=FALSE, warning=FALSE}
library(tidyverse)
library(openintro)
library(gghighlight)
library(latex2exp)
data(COL)
seed <- 42
```

## Objectives

:::: {.column width=15%}
::::

:::: {.column width=70%}
- **Develop and understanding of the exponential distribution**
- **Develop and understanding of the normal distribution**
- **Know how to compute exponential and normal probabilities using R**
- **Introduce how to simulate exponential and normal random sampling in R**
::::

:::: {.column width=15%}
::::

## Visualizing The Exponential Distribution (1/2)

:::: {.column width=50%}
**Exponential Probability Density Function (PDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# exponential pdf
mean <- 3
lambda <- 1/mean
x_exp <- seq(0,20,0.10)
exp_pdf <- dexp(x_exp,lambda)

# convert pdf to tibble
df_exp <- tibble(x=x_exp,pdf=exp_pdf)

# plot the exponential distribution
ggplot(df_exp,aes(x=x,y=pdf)) + 
  geom_line(color="black",linewidth=1) + 
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal()
```

::: {style="color: red;"}
$\star$ The exponential r.v. has infinitely sized sample space. The plot of the exponential PDF only shows the range $0 \le x \le 20$.
:::
::::

:::: {.column width=49%}
**Exponential R.V. and PDF:**

Let $\lambda$ be the rate parameter.

* $\displaystyle X \sim \text{Exp}\left(\lambda\right)$
* $\displaystyle f(x) = \lambda e^{-\lambda x}, \quad x \ge 0$

**Relevant R Functions:**

* `dexp` $\leftarrow$ PDF
* `rexp` $\leftarrow$ random sampling simulation
::::

## Visualizing The Exponential Distribution (2/2)

:::: {.column width=50%}
**Exponential Cumulative Density Function (CDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=2,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 3
lambda <- 1/mean
x_exp <- seq(0,20,0.10)
exp_cdf <- pexp(x_exp,lambda)

# convert cdf to tibble
df_exp <- tibble(x=x_exp,cdf=exp_cdf)

# plot the exponential distribution
ggplot(df_exp,aes(x=x,y=cdf)) + 
  geom_line(color="black",linewidth=1) + 
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal()
```

::: {style="color: red;"}
$\star$ The CDF, in general, computes the cumulative probability of a given PDF.
:::
::::

:::: {.column width=49%}
**Exponential PDF and CDF:**

Let $\lambda$ be the rate parameter.

* $\displaystyle f(x) = \lambda e^{-\lambda x}, \quad x \ge 0$
* $\displaystyle P(X \le x) = \int_{0}^{x} f(t) \ dt, \quad x \ge 0$

**Relevant R Functions:**

* `pexp` $\leftarrow$ CDF
* `qexp` $\leftarrow$ Inverse CDF
::::

## Exponential Probabilities (1/2)

:::: {.column width=50%}
**Exponential Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 3
lambda <- 1/mean 
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 60 (technically infinite)
exp_pdf <- dexp(x_outcomes,lambda) # function `dexp()` is the Exponential PDF 
df <- tibble(x=x_outcomes, pmf=exp_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pmf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=0 & x<=2.5),aes(x=x,ymax=dexp(x,lambda)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Note that the `pexp` function computes the probability $P(X \le x)$, meaning it computes the area under the PDF from $X=0$ to $X=x$. The `dexp` function computes the density, not probability because $P(X = x) = 0$ at any $x$.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that "success" happens on 2.5 unit length or less, given $\lambda=\frac{1}{3}$? *
\[
\begin{aligned}
P(X \le 2.5) & = \int_0^{2.5} f(t) \ dt \\
            & = \int_0^{2.5} \frac{1}{3} e^{-\frac{1}{3} t} \ dt \\
P(X \le 2.5) & \approx 0.565
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
pexp(2.5,1/3)
```
::::

## Exponential Probabilities (2/2)

:::: {.column width=50%}
**Exponential Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 3
lambda <- 1/mean 
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 60 (technically infinite)
exp_pdf <- dexp(x_outcomes,lambda) # function `dexp()` is the Exponential PDF 
df <- tibble(x=x_outcomes, pmf=exp_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pmf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=0 & x>=2.5),aes(x=x,ymax=dexp(x,lambda)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Use the complement rule when computing the probability $P(X \ge x) = 1 - P(X \le x)$.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that "success" happens on at least 2.5 unit length, given $\lambda=\frac{1}{3}$? *
\[
\begin{aligned}
P(X \ge 2.5) & = 1 - P(X \le 2.5) \\
            & = 1 - \int_0^{2.5} f(t) \ dt \\
            & = 1 - \int_0^{2.5} \frac{1}{3} e^{-\frac{1}{3} t} \ dt \\
            & = \int_{2.5}^{\infty} \frac{1}{3} e^{-\frac{1}{3} t} \ dt \\
P(X \ge 2.5) & \approx 0.435
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
1-pexp(2.5,1/3)
```
::::

## Exponential Interval Probabilities

:::: {.column width=50%}
**Exponential Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 3
lambda <- 1/mean 
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 60 (technically infinite)
exp_pdf <- dexp(x_outcomes,lambda) # function `dexp()` is the Exponential PDF 
df <- tibble(x=x_outcomes, pmf=exp_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pmf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=1 & x<=5),aes(x=x,ymax=dexp(x,lambda)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Here, you need $P(1 \le X \le 5)$, which is equivalent to $P(X \le 5) - P(X \le 1)$. Since $P(X = 1) = 0$ and $P(X = 5) = 0$ for continuous r.v.s, we don't worry about these cases when calculating interval probabilities.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that "success" happens between 1 to 5 unit length, given $\lambda=\frac{1}{3}$? *
\[
\begin{aligned}
P(1 \le X \le 5) & = P(X \le 5) - P(X \le 1) \\
            & = \int_0^{5} f(t) \ dt - \int_0^{1} f(t) \ dt \\
            & = \int_1^{5} f(t) \ dt \\
            & = \int_1^{5} \frac{1}{3} e^{-\frac{1}{3} t} \ dt \\
P(1 \le X \le 5) & \approx 0.528
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
pexp(5,1/3)-pexp(1,1/3)
```
::::

## Exponential Expected Value

:::: {.column width=50%}
**Exponential Distribution with Expected Value**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mean <- 3
lambda <- 1/mean
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 60 (technically infinite)
exp_pdf <- dexp(x_outcomes,lambda) # function `dexp()` is the Exponential PDF 
df <- tibble(x=x_outcomes, pdf=exp_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_vline(xintercept = mean, color="red", linewidth=1) +
  geom_text(aes(mean,0),label = paste(TeX("E(X)")," = ",mean,sep=""), 
            vjust = -8, hjust=-0.10,color="red") + 
  ggtitle(TeX(paste("Exponential PDF ","$(\\lambda","=","1/",mean,")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ The law of large numbers still can be used to interpret the expected value, which in this case, you would expect to have $3$ unit length until a "success" occurs in the long run.
:::
::::

:::: {.column width=49%}
**Exponential R.V.**

Let $\lambda=\frac{1}{3}$ be the rate parameter.

* $\displaystyle X \sim \text{Exp}\left(\frac{1}{3}\right)$
* $\displaystyle f(x) = \lambda e^{-\frac{1}{3} x}, \quad x \ge 0$
* $\displaystyle \text{E}(X) = 3$

In general, 
\[
\begin{aligned}
\text{E}(X) & = \int_0^{\infty} x f(x) \ dx \\
            & = \int_0^{\infty} x \lambda e^{-\lambda x} \ dx \\
\text{E}(X) & = \frac{1}{\lambda}
\end{aligned}
\]
which is the reciprocal of the rate parameter.
::::

## Visualizing the Normal Distribution (1/2)

:::: {.column width=50%}
**Normal Probability Density Function (PDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_pdf <- dnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, pdf=norm_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ The normal r.v. has infinitely sized sample space. The plot of the exponential PDF only shows the range $0 \le x \le 20$.
:::
::::

:::: {.column width=49%}
**Normal R.V. and PDF:**

Let $\mu$ be the mean and $\sigma^2$ be the variance.

* $\displaystyle X \sim \text{N}\left(\mu,\sigma^2\right)$
* $\displaystyle f(x) = \frac{1}{\sigma \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{x - \mu}{\sigma} \right)^2\right)}, \ -\infty < x < \infty$

**Relevant R Functions:**

* `dnorm` $\leftarrow$ PDF
* `rnorm` $\leftarrow$ random sampling simulation
::::

## Visualizing the Normal Distribution (2/2)

:::: {.column width=50%}
**Normal Cumulative Density Function (CDF)**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=4,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_cdf <- pnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, cdf=norm_cdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=cdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ The CDF of the Normal distribution is also known as the *sigmoid* function.
:::
::::

:::: {.column width=49%}
**Normal PDF and CDF:**

Let $\mu$ be the mean and $\sigma^2$ be the variance.

* $\displaystyle f(x) = \frac{1}{\sigma \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{x - \mu}{\sigma} \right)^2\right)}, \ -\infty < x < \infty$
* $\displaystyle P(X \le x) = \int_{\infty}^{x} f(t) \ dt$

**Relevant R Functions:**

* `pnorm` $\leftarrow$ CDF
* `qnorm` $\leftarrow$ Inverse CDF
::::

## Normal Probabilities (1/2)

:::: {.column width=50%}
**Normal Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=5,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_pdf <- dnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, pdf=norm_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=0 & x<=13),aes(x=x,ymax=dnorm(x,mu,sd)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Note that the `pnorm` function id the CDF, which computes the probability $P(X \le x)$, meaning it computes the area under $f(x)$, the PDF, from $-\infty$ to $X=x$. The `dnorm` function computes the density, not probability because $P(X = x)=0$ at any $x$.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that a randomly selected value is at most $13$?*

\[
\begin{aligned}
P(X \le 13) & = \int_{-\infty}^{13} f(t) \ dt \\
            & = \int_{-\infty}^{13} \frac{1}{2.24 \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{t - 10}{2.24} \right)^2\right)} \ dt \\
P(X \le 13) & \approx 0.910
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
pnorm(13,10,2.24)
```
::::

## Normal Probabilities (2/2)

:::: {.column width=50%}
**Normal Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=5,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_pdf <- dnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, pdf=norm_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=0 & x>=13),aes(x=x,ymax=dnorm(x,mu,sd)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Remember that there is no known anti-derivative to the normal PDF (also known as the Gaussian function). So, the integral here is evaluated using numerical approximations.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that a randomly selected value is at least $13$?*

\[
\begin{aligned}
P(X \ge 13) & = 1 - P(X \le 13) \\
            & = 1 - \int_{-\infty}^{13} f(t) \ dt \\
            & = 1 - \int_{-\infty}^{13} \frac{1}{2.24 \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{t - 10}{2.24} \right)^2\right)} \ dt \\
            & = \int_{13}^{\infty} \frac{1}{2.24 \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{t - 10}{2.24} \right)^2\right)} \ dt \\
P(X \le 13) & \approx 0.090
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
1-pnorm(13,10,2.24)
```
::::

## Normal Interval Probabilities

:::: {.column width=50%}
**Normal Distribution:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=5,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_pdf <- dnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, pdf=norm_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_ribbon(data=subset(df,x>=7 & x<=13),aes(x=x,ymax=dnorm(x,mu,sd)),ymin=0,alpha=0.3,fill="blue") +
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```

::: {style="color: red;"}
$\star$ Here, you need $P(7 \le X \le 13)$, which is equivalent to $P(X \le 13) - P(X \le 7)$. Since $P(X = 7) = 0$ and $P(X = 13) = 0$ for continuous r.v.s, we don't worry about these cases when calculating interval probabilities.
:::
::::

:::: {.column width=49%}
**Example:**

*What is the probability that a randomly selected value is between $7$ and $13$?*

\[
\begin{aligned}
P(7 \le X \le 13) & = P(X \le 13)-P(X \le 7) \\
            & = \int_{-\infty}^{13} f(t) \ dt - \int_{-\infty}^{7} f(t) \ dt \\
            & = \int_{7}^{13} f(t) \ dt \\
            & = \int_{7}^{13} \frac{1}{2.24 \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{t - 10}{2.24} \right)^2\right)} \ dt \\
P(7 \le X \le 13) & \approx 0.820
\end{aligned}
\]

**Using R:**

```{r echo=TRUE, eval=TRUE}
pnorm(13,10,2.24)-pnorm(7,10,2.24)
```
::::

## Normal Expected Value

:::: {.column width=50%}
**Normal Distribution with Expected Value:**

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=5,fig.height=3,out.width='100%'}
# set parameters, outcomes, and probabilities
mu <- 20*0.50
sd <- sqrt(20*0.50*(1-0.50))
x_outcomes <- seq(0,20,0.10) # set outcome from 0 to 20 (technically infinite)
norm_pdf <- dnorm(x_outcomes,mu,sd) # function `dnorm()` is the Normal PDF 
df <- tibble(x=x_outcomes, pdf=norm_pdf) # convert to tibble

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df,aes(x=x,y=pdf)) + 
  geom_line(size=1,color="black") + # size here is defined for all points
  geom_vline(xintercept = mu, color="red", linewidth=1) + 
  geom_text(aes(mu,0),label = paste(TeX("E(X)")," = ",mu,sep=""), 
            vjust = -1, hjust=-0.10,color="red") + 
  ggtitle(TeX(paste("Normal Distribution ","$(\\mu","=",mu," ",", \\sigma","=",round(sd,2),")$"))) + # sets the title of the plot
  theme_minimal() # set theme of entire plot

# display plot
p1
```
::::

:::: {.column width=49%}
**Exponential R.V.**

Let $\mu = 10$ be the mean and $\sigma^2 = 2.24^2$ be the variance.

* $\displaystyle X \sim \text{N}\left(10,2.24^2\right)$
* $\displaystyle f(x) = \frac{1}{\sigma \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{x - \mu}{\sigma} \right)^2\right)}, \ -\infty < x < \infty$
* $\displaystyle \text{E}(X) = 10$

In general, the expected value of the normal r.v. is given by 
\[
\begin{aligned}
\text{E}(X) & = \int_{-\infty}^{\infty} x f(x) \ dx \\
            & = \int_{-\infty}^{\infty} \frac{x}{\sigma \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{x - \mu}{\sigma} \right)^2\right)} \ dx \\
\text{E}(X) & = \mu
\end{aligned}
\]
which is the center of the normal distribution.
::::

## The Normal Distribution is Symmetric

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(x=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=x,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) +
  geom_vline(xintercept=0, color="black") + 
  ylab("density") + 
  ggtitle("Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-3,-2,-1,0,1,2,3),label=c(TeX("$\\mu-3\\sigma$"),TeX("$\\mu-2\\sigma$"),TeX("$\\mu-1\\sigma$"),TeX("$\\mu$"),TeX("$\\mu+1\\sigma$"),TeX("$\\mu+2\\sigma$"),TeX("$\\mu+3\\sigma$"))) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

::: {style="color: red;"}
$\star$ The key property of the normal distribution is that it is symmetric about the mean $\mu$, meaning the probabilities are equal for each side.
:::

## The 68-95-99.7 Rule (1/3)

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(x=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=x,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_vline(xintercept=1, color="black") + 
  geom_vline(xintercept=-1, color="black") + 
  geom_ribbon(data=subset(df_norm,x>=-1 & x<=1),aes(x=x,ymax=dnorm(x,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-3,-2,-1,0,1,2,3),label=c(TeX("$\\mu-3\\sigma$"),TeX("$\\mu-2\\sigma$"),TeX("$\\mu-1\\sigma$"),TeX("$\\mu$"),TeX("$\\mu+1\\sigma$"),TeX("$\\mu+2\\sigma$"),TeX("$\\mu+3\\sigma$"))) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

$$P(\mu - \sigma \le X \le \mu + \sigma) \approx 0.68$$

## The 68-95-99.7 Rule (2/3)

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(x=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=x,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_vline(xintercept=2, color="black") + 
  geom_vline(xintercept=-2, color="black") + 
  geom_ribbon(data=subset(df_norm,x>=-2 & x<=2),aes(x=x,ymax=dnorm(x,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-3,-2,-1,0,1,2,3),label=c(TeX("$\\mu-3\\sigma$"),TeX("$\\mu-2\\sigma$"),TeX("$\\mu-1\\sigma$"),TeX("$\\mu$"),TeX("$\\mu+1\\sigma$"),TeX("$\\mu+2\\sigma$"),TeX("$\\mu+3\\sigma$"))) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

$$P(\mu - 2\sigma \le X \le \mu + 2\sigma) \approx 0.95$$

## The 68-95-99.7 Rule (3/3)

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(x=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=x,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_vline(xintercept=3, color="black") + 
  geom_vline(xintercept=-3, color="black") + 
  geom_ribbon(data=subset(df_norm,x>=-3 & x<=3),aes(x=x,ymax=dnorm(x,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-3,-2,-1,0,1,2,3),label=c(TeX("$\\mu-3\\sigma$"),TeX("$\\mu-2\\sigma$"),TeX("$\\mu-1\\sigma$"),TeX("$\\mu$"),TeX("$\\mu+1\\sigma$"),TeX("$\\mu+2\\sigma$"),TeX("$\\mu+3\\sigma$"))) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

$$P(\mu - 3\sigma \le X \le \mu + 3\sigma) \approx 0.997$$

## Total Area

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(x=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=x,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_ribbon(data=subset(df_norm,x>=-4 & x<=4),aes(x=x,ymax=dnorm(x,mu,sigma)),ymin=0,alpha=0.3,fill="#009159") +
  ylab("density") + 
  ggtitle("Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-3,-2,-1,0,1,2,3),label=c(TeX("$\\mu-3\\sigma$"),TeX("$\\mu-2\\sigma$"),TeX("$\\mu-1\\sigma$"),TeX("$\\mu$"),TeX("$\\mu+1\\sigma$"),TeX("$\\mu+2\\sigma$"),TeX("$\\mu+3\\sigma$"))) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

$$P(\mu - \infty \le X \le \mu + \infty) = 1$$

::: {style="color: red;"}
$\star$ Because of the axiom that the sum of the probabilities for all outcomes in the sample space is equal to 1, the total area under the Normal PDF is always 1.
:::

## The Standard Normal Distribution

:::: {.column width=49%}
**The z-score formula:**

$$z = \frac{x - \mu}{\sigma}$$

**Standard Normal R.V. and PDF:**

Let $\mu = 0$ be the mean and $\sigma^2 = 1$ be the variance.

* $\displaystyle Z \sim \text{N}\left(0,1\right)$
* $\displaystyle f(z) = \frac{1}{\sqrt{2 \pi}} \exp{\left(-\frac{z^2}{2} \right)}; \\ \quad \text{ for } -\infty < z < \infty$
::::

:::: {.column width=49%}
**How the transformation works:**

* Let $x = \mu$ and $\sigma = 1$.
* Then, $z = 0$, meaning the distance from the mean to itself is $0$ standard deviations.
* A positive z-score indicates the value is above the mean, while a negative z-score indicates it is below.

::: {style="color: red;"}
$\star$ The z-score $z$ is a measure on how many standard deviations a data point $x$ is from the mean $\mu$.
:::
::::

## Visualizing the Standard Normal Distribution

```{r eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center',fig.width=7,fig.height=3,out.width='80%'}
# normal pdf
mu <- 0
sigma <- 1
x_norm <- seq(-4,4,0.10)
norm_pdf <- dnorm(x_norm,mu,sigma)

# convert pdf into tibble
df_norm <- tibble(z=x_norm, norm_pdf=norm_pdf)

# plot the Bernoulli distribution and store it into a R variable
p1 <- ggplot(df_norm,aes(x=z,y=norm_pdf)) + 
  geom_line(color="#009159",linewidth=1) + 
  geom_vline(xintercept=0, color="black") + 
  ylab("density") + 
  ggtitle("Standard Normal Distribution") + # sets the title of the plot
  scale_x_discrete(limits=c(-4,-3,-2,-1,0,1,2,3,4)) + 
  theme_minimal() + # set theme of entire plot
  theme(legend.title=element_blank())

# display plot
p1
```

::: {style="color: red;"}
$\star$ The standard normal distribution serves as a reference distribution, allowing any normally distributed variable to be standardized.
:::

## Summary of Continuous R.V.s and their Properties

|  | **Exponential** | **Normal** |
|:----|:-------:|:-------:|
| *Description* | Unit length until a "success" event happens | Approximation to the Binomial with sufficiently large number of $n$ independent Bernoulli trials |
| *Parameters* | $\lambda \longrightarrow$ rate of "success" | $\mu \longrightarrow$ mean<br>$\sigma^2 \longrightarrow$ variance |
| $\displaystyle X$ | $\displaystyle \text{Exp}(\lambda)$ | $\displaystyle \text{N}(\mu,\sigma^2)$ |
| $\displaystyle \text{E}(X)$ | $\displaystyle \frac{1}{\lambda}$ | $\displaystyle \mu$ |
| $\displaystyle \text{Var}(X)$ | $\displaystyle \frac{1}{\lambda^2}$ | $\displaystyle \sigma^2$ |

::: {style="color: red;"}
$\star$ If you are approximating a Binomial distribution, set $\mu = np$ and $\sigma^2 = np(1-p)$.

$\star$ If you want the standard normal distribution set $\mu = 0$ and $\sigma = 1$.
:::

## Summary of Continuous R.V.s and their PMFs

|  | **Exponential** | **Normal** |
|:----|:-------:|:-------:|
| $\displaystyle f(x)$ *PDF* | $\displaystyle \lambda e^{-\lambda x}$<br>$\text{for } x \ge 0$ | $\displaystyle \frac{1}{\sigma \sqrt{2 \pi}} \exp{\left(-\frac{1}{2} \left( \frac{x - \mu}{\sigma} \right)^2\right)}$<br>$\text{for }-\infty < x < \infty$ |
| *R PMF* | `dexp` | `dnorm` |
| *R CDF* | `pexp` | `pnorm` |
| *R Inverse CDF* | `qexp` | `qnorm` |
| *R Simulations* | `rexp` | `rnorm` |
