---
title: "Data Sub-Setting and Identifying Variables"
subtitle: "Applied Statistics"
author: "<span style='color:blue'>Solutions</span>"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Syllabus Reminders

- You must submit your worksheet individually by end-of-class or end-of-day. Your name must exist in your worksheet and the names of your collaborators.
- Worksheets are marked mostly on completion, and partially on correctness. It will be marked either pass or fail, there will no detailed feedback on worksheets, and no opportunities for revisions and make-up.

# Instructions

- For the exercises, please provide your answers in full sentences.
- Write your answers by replacing the text "[Write your answer here]".
- Whenever a problem asks you to run something in R, create an R code chunk and write your R commands inside that chunk.
- If a problem requires modifying an existing R code chunk, make the necessary changes at the appropriate lines within the given code.
- Please read the problems fully and carefully before answering.
- When finished, knit this `.Rmd` to `.html`.
- Submit this `.Rmd` and the recently knitted `.html` to Moodle.

# Packages

The following R packages have already been loaded for you, so there's no need to load them yourself—they will be used in the problem sets. If you need any additional packages, you can load them here as well.

```{r message=FALSE, warning=FALSE}
# load packages
library(tidyverse)
```

# Collaborators

::: {style="color: blue;"}
"<span style='color:blue'>Solutions</span>"
:::

# Exercises

## Problem 1

The following code chunk installs the [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/){target="_blank"} package, which contains the `penguins` data set. The R code chunk has an option called `eval=FALSE`, which indicate that --when knitting-- the code chunk will not run. Explain what happens after a knitting attempt if `eval=TRUE`.

```{r install-palmerpenguins, eval=FALSE}
install.packages("palmerpenguins")
```

::: {style="color: blue;"}
The running the above code chunk within the RMarkdown file, the code within the chunk is run using the console. When knitting the entire RMarkdown the said code chunk is displayed but not it was not executed.
:::

## Problem 2

Create a code chunk. In this chunk, load the `palmerpenguins` package. Are there any warning messages after loading the packages? If so, what does it say and what do you think it means?

::: {style="color: blue;"}
```{r}
library(palmerpenguins)
```

When running the above code chunk with `library(palmerpenguins)`, there was some warning output message and it says "Warning: package ‘palmerpenguins’ was built under R version 4.4.2" and it means that the `palmerpenguins package was made using the older version of R.
:::

## Problem 3

Create another code chunk. In this chunk, show a preview of the `penguins` data frame. How many rows and columns are there in the `penguins` data set? Is the data frame a base R data frame or a tibble? Explain your reasoning.

::: {style="color: blue;"}
```{r}
penguins
```

There are 344 rows and 8 columns. The data frame is in the `tibble` format.
:::

## Problem 4

Statistical variables are classified into two main types: numerical and categorical. Numerical variables can be further divided into discrete and continuous subtypes, while categorical variables can be classified as nominal or ordinal.

Each row of the `penguins` data set is an observation of a penguin. Use your output in the previous problem to answer the following question: What type of statistical variables are there in the columns of the data frame? Explain your reasoning.

::: {style="color: blue;"}
* `species` | nominal categorical | The categories within this variable appears to have no discernible order.
* `island` | nominal categorical | The same reasoning as the `species` variable.
* `bill_length_mm` | continuous numerical | This is a measurable variable.
* `bill_depth_mm` | continuous numerical | The same as `bill_length_mm`.
* `flipper_length_mm` | continuous numerical | It appears to be discrete but this is a measurable variable.
* `body_mass_g` | continuous numerical | The same reasoning as the `flipper_length_mm`.
* `sex` | nominal categorical | The categories within this variable appears to have no discernible order.
* `year` | discrete numerical or ordinal categorical | This appears to be a discrete numerical but it is a time-based variable. So, `year` can also be an ordinal variable.
<dbl>

:::

## Problem 5

Subset the `penguins` data set using the following four cases. Each case should be a new tibble in a new R variable. Display the resulting tibbles using the `glimpse()` function.

* All rows with the "Adelie" species
* All rows in the year 2008
* All rows with bill length at least $40$ mm.
* All rows with the "Adelie" species in the year 2008, and with a bill length at least $40$ mm.

Make sure to write your R code clearly in a code chunk and include comments explaining the purpose of each step.

::: {style="color: blue;"}
```{r}
# All rows with the "Adelie" species
sub1 <- penguins[penguins$species == "Adelie",]
glimpse(sub1)

# All rows in the year 2008
sub2 <- penguins[penguins$year == 2008,]
glimpse(sub2)

# All rows with bill length at least $40$ mm.
sub3 <- penguins[penguins$bill_length_mm >= 40,]
glimpse(sub3)

# All rows with the "Adelie" species in the year 2008, and with a bill length at least $40$ mm.
sub4 <- sub1[sub1$year == 2008,]
sub5 <- sub4[sub4$bill_length_mm >= 40,]
glimpse(sub5)
```
:::
